State-space PTA

### 2023-04-02 docs/manuscript/paper.tex
assorted runs

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, O'Leary, Melatos, Evans, others, etc. %
\thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}

@@

@@

\section{Model}
We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \cite{Vargas}. The frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proporionality constant, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
Over the timescales that we are interested in, we can express the EM spindown simply as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
This intrinsic frequency can be related to a measured frequency as 
\begin{equation}
	f_M = f g(\bar{\theta},t) + N_M
\end{equation}
where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
\begin{equation}
	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
\end{equation}
Explicitly, it can be shown that the measurement function is,
\begin{equation} \label{eq:final}
	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q(t)^i q(t)^j}{1 + \bar{n} \cdot \bar{q}(t)}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}(t)) d}]
\end{equation}
There are a few terms to unpack and define in this equation. $q(t)$ is the vector from the Earth to the pulsar, $n$ the propagation direction of the GW, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance. $h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$. It is given by the familiar plane-wave relation:	
\begin{equation}
	h_{ij}(t) = H_{ij} e^{(i(\bar{k} \cdot \bar{x} - \Omega t + \Phi_0))}
\end{equation}
where $\bar{k}$ is the GW 3-wavevector $=\Omega \bar{n}$ and $x = -\bar{q} t$ (photon propagates from pulsar to Earth), and for phase offset (GW phase at Earth when $t=0$) $\Phi_0$.  We can also write this as,
\begin{equation}
h_{ij}(t) = H_{ij} e^{(-i\Omega t (1 + \bar{n} \cdot \bar{q} ) + \Phi_0))}
\end{equation}
The amplitude tensor  $H_{ij}$ is
\begin{equation}
	H_{ij} = h_+ e_{ij}^+(\bar{n},\psi) + h_{\times} e_{ij}^{\times}(\bar{n},\psi)
\end{equation}
where $h_{+,\times}$ are the constant amplitudes of the gravitational plane wave, and $e_{ij}^{+, \times}(\bar{n}, \psi)$ are the polarisation tensors which are uniquely defined by the principal axes of the GW. $\psi$ is the polarisation angle of the GW. \newline

@@

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, Melatos,O'Leary, , Evans, others, etc. %
\thanks{Contact e-mail: \href{mailto:mn@ras.ac.uk}{mn@ras.ac.uk}}%
\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}

@@

The detection of high frequency ($\sim 1$Hz) gravitational waves (GWs) from coalescing BH binaries with ground-based detectors such as LIGO/Virgo \citep{aLIGO,2015CQGra..32b4001A} is now a routine enterprise \citep[e.g.][]{2019PhRvX...9c1040A,2021PhRvX..11b1053A}. Gravitational radiation from sources which radiate in the mill-Hz regime are expected to be detectable from $\sim 2030$ with the space-based Laser Interferometer Space Antenna, \citep{LISApaper} especially given the early success by the pathfinder mission \citep{2019arXiv190308924A}. Detecting GWs from systems which evolve over even longer timescales, $\mathcal{O}$(years), has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this nano-Hz regime is a via timing an ensemble of milliseconds pulsars; a pulsar timing array (PTA) \citep{2021hgwa.bookE...4V}. The presence  of a nano-Hz gravitational wave will influence the propagation of the pulsar radio beacon and consequently will imprint on the pulsar timing signal. By measuring the modulation of the received pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. \newline

\noindent Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, along with a number of newer PTAs, under the umbrella of the International Pulsar Timing Array \citep[IPTA][]{2019MNRAS.490.4666P}. The primary target of PTA observations is the inspiral of supermassive black hole binaries (SMBHBs) with masses $\sim 10^7 M_{\odot}$. These signals from SMBHBs can be broadly classified into either deterministic or stochastic. For the former, sufficiently bright and near binaries may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated. For the latter, the  incoherent superposition of multiple weaker SMBHBs sources leads to a stochastic background detectable at nano-Hz frequencies. Other potential sources for PTAs include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}, but the deterministic and stochastic GW signals from SMBHBs remain the primary targets. \newline

\noindent The detection of loud, resolved sources with a PTA typically involves a parametrised model for the pulsar timing residuals induced by the modulation of the pulsar signal by a GW. One can then search for evidence that this model describes the data via the usual Bayesian likelihood techniques, and try to estimate the parameters of the model \citep[e.g.][]{Babak2016}. For detecting the stochastic background the approach is different; one measures the correlation in pulsar timing residuals between any two pair of pulsars. The presence of a GW induces a characteristic correlation function as a function of the angular separation between the pulsars; the Hellings-Downs curve \citep{Hellings}. For both classes of source, detection of GW signals in the timing residuals of a PTA is a challenging enterprise, and currently neither a stochastic background nor an individually resolved source has yet been detected \citep{2022MNRAS.510.4873A, 10.1093/nsr/nwx126}. \newline

\noindent Motivated by the difficulties faced by the classic PTA analysis methods, in this work we present a novel approach to formulate PTA analysis and GW detection as a state-space problem. This approach enables the state-space evolution to be tracked optimally, and for a specific realisation of the pulsar process noise (i.e. the spin wandering of the pulsar) infer both the system parameters and the detectability of the GW signal.

For this initial work we will focus exclusively on resolved, monochromatic GW sources. Discussion on

This paper is organised as follows.

In section 2 we will briefly outline the response 
In section 3 we will then construct a state space form for determinsitic signals (3.1) outline the mehtods for expectaiton likelihood (3.2) and apply this state space kalma nform

In section 4 we will extend to stochastic sources...? OR we will leave thus for a second paper...

PTA sources can be braodly characterised into two main buckets: determinsitic and stocahstic. Derterminisitc sources are loud isngle

@@

\section{Evolution of the pulsar frequency}
We will take as our model of the intrinsic pulsar frequency $f$ a variation of the phenomenological model of \citep{Vargas}. Within this model, $f$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering', achromatic variations in the pulse TOA intrinsic to the star). Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\gamma$ a proportionality constant, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
For PTA analysis, we are concerned with timescales on the order or years. Consequently, we can express the EM spindown straightforwardly as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df}{dt} = -\gamma	 [f - f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}

\section{Modulation of pulsar frequency due to a GW}
In the presence of a GW, the $f(t)$ measured by an observer on Earth is different from that measured by an observer in the local NS reference frame. We want to determine how the GW influences the received frequency

\subsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
for spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that $\Omega$ has no time dependence - we are concerned solely with monochromatic sources. Note that we are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ \textit{at the Earth.}\footnote{This point is important, since the phase offset is then the same between multiple pulsars. See also Melatos 2022 PT12}. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
\begin{equation}
	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
\end{equation}
\begin{equation}
		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
\end{equation}
which are in turn specified via the location of the GW source on the sky (via $\theta$, $\phi$ coordinates) and the polarisation angle $\psi$

\begin{align}
	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}

\subsection{Pulse frequency as a photon}
\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon with 4-momentum $p_{\mu}$ recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	\nu = p_{\alpha} u^{\alpha}
\end{equation}
\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the frequency can be identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	\nu = p_t
\end{equation}
\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
\end{equation}
\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path. \footnote{\textbf{TK: See also e.g. Maggiore, Meltaos who takes the same approach...}} \newline

\noindent To complete our expression, we now just need to determine $\dot{p}_t$ and integrate it.

\subsection{Hamiltonian Mechanics}
The Hamiltonian in covariant notation can be written as 
\begin{equation}
	H(x^{\mu}, p_{\mu}) = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
\end{equation}
\noindent which if we substitute in our expression for the perturbed metric is
\begin{equation}
	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
\noindent Now, Hamilton's equations are
\begin{equation}
	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
\end{equation}

\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
\begin{equation}
	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t))+\Phi_0}
\end{equation}
\footnote{This is equivalent to Melatos 2018, Eq 5 for the specific case of a GW propagating in the $z$-direction, with zero phase offset.}

\noindent Therefore the derivative w.r.t coordinate time $t$ is,
\begin{equation}
	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
\end{equation}
\noindent To make contact with Melatos 2022 it will prove useful to recognise that $p^{\mu} = \omega(1,-q^x,-q^y,-q^z)$ where $\bar{q}$ is the unit vector between the Earth and pulsar \footnote{similarly $\bar{x}(t) = -\bar{q}t$} and $\omega$ is the \textit{constant} photon angular frequency. Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer (see Melatos 2022, PT16). \newline

\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic.  \newline

\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
\begin{equation}
	\dot{p}_t = A e^{i \gamma t + \Phi_0}
\end{equation}
where 
\begin{equation}
	\gamma = -\Omega (1 + \bar{n}\cdot \bar{q}) 
\end{equation}
\footnote{compare with Melatos 2022, PT11}
\noindent and
\begin{equation}
	A = -\frac{i\Omega \omega}{2} H_{ij}q^i q^j 
\end{equation}
where the $\omega$ term in the numerator results from a $\omega^2$ terms that comes from the $p^i p^j$ term and a $\omega^{-1}$ term from $(p^t)^{-1}$.
\subsection{Performing the integral}

The frequency shift experienced by the observer relative to the source due to a GW is then
\begin{align}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} = A \int_{t = \tau - d}^{t=\tau} e^{i \gamma t + \Phi_0}dt  \\
	=\frac{-iA}{\gamma} e^{i \gamma \tau}e^{\Phi_0}[1 - e^{-i \gamma d}] 
	\label{eq:concise}
\end{align}

\subsection{Explicit expression and comparison with Melatos 22}

We can expand the concise expression of Eq. \ref{eq:concise} as,

\begin{equation} \label{eq:final}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{H_{ij}q^i q^j  }{(1 + \hat{n} \cdot \hat{q})}e^{-i \Omega \tau(1+\bar{n} \cdot \bar{q}) + \Phi_0} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}
This expression is equivalent to Melatos 2022, PT13 where the dependence on time $t$ is made explicit. Using the Melatos notation where $h_{ij} = g_{ij} - \eta_{ij}$

\begin{equation}
	p_t(\tau)|_{\rm Earth} - p_t(\tau - d)|_{\rm source} =\frac{ \omega}{2} \frac{h_{ij} q^i q^j  }{(1 + \hat{n} \cdot \hat{q})} [1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}) d}]
\end{equation}

\newpage
\section{State-space structure} \label{sec3}
\noindent The general structure of a Kalman state-space problem is
\begin{equation}
	\dot{x} = f(x) + w
\end{equation}
where $x$ are the state variables, $f()$ a non-linear function of the states, and $w$ a stochastic zero-mean process. The process noise matrix $Q = E(w w^T)$. The states are related to the measurement $z$ via a non-linear measurement function $h()$
\begin{equation}
	z = h(x) + v
\end{equation}
where $v$ is a stochastic zero-mean process and measurement noise matrix $R = E(v v^T)$. \newline

\noindent This structure maps onto the preceding equations nicely. Our hidden state is just the intrinsic pulsar frequency, which evolves as,
\begin{eqnarray}
	\dot{f}_P = -\gamma f_P^n + \xi(t)
\end{eqnarray}
where $n$ is a braking index (typical values $\sim 1-5$, we take a canonical value $=3$) \footnote{See A. Vargas talk re anomalous braking indices. P.S. is there an arXiv preprint?} $\gamma$ a proportionality constant, and $\xi(t)$ represents a stochastic white noise process. Equation \ref{eq:final} can be used to relate the intrinsic pulsar frequency to that measured by an observer on Earth:

\begin{eqnarray}
	f_M = f_p(1 - X) + N_M
\end{eqnarray}
where $X$ is the RHS of Eq. \ref{eq:final} and $N_M$ a Gaussian measurement noise. An example of the evolution of the state and the measurement frequencies for our PTA pulsars can be seen in Fig. \ref{fig:mean and std of nets}

\section{Discretisation}

This intrinsic frequency can be related to a measured frequency as 
\begin{equation}
	f_M = f g(\bar{\theta},t) + N_M
\end{equation}
where $g(\theta,t)$ ("measurement function")is a function of some parameters, $\bar{\theta}$ and time $t$, whilst $N_M$ is a Gaussian measurement noise that satisfies 
\begin{equation}
	\langle N_M(t) N_M(t') \rangle = \Sigma^2 \delta(t - t')
\end{equation}
Explicitly, it can be shown that the measurement function is,
\begin{equation} \label{eq:final}
	g(\bar{\theta},t) = 1 - \frac{1}{2} \frac{h_{ij}(t) q(t)^i q(t)^j}{1 + \bar{n} \cdot \bar{q}(t)}[1 - e^{i \Omega(1+\bar{n} \cdot \bar{q}(t)) d}]
\end{equation}
There are a few terms to unpack and define in this equation. $q(t)$ is the vector from the Earth to the pulsar, $n$ the propagation direction of the GW, $\Omega$ the (constant) angular frequency of the GW and $d$ the Earth-pulsar distance. $h_{ij}(t)$ is the metric perturbation due to the GW $= g_{ij} - \eta_{ij}$. It is given by the familiar plane-wave relation:	
\begin{equation}
	h_{ij}(t) = H_{ij} e^{(i(\bar{k} \cdot \bar{x} - \Omega t + \Phi_0))}
\end{equation}
where $\bar{k}$ is the GW 3-wavevector $=\Omega \bar{n}$ and $x = -\bar{q} t$ (photon propagates from pulsar to Earth), and for phase offset (GW phase at Earth when $t=0$) $\Phi_0$.  We can also write this as,
\begin{equation}
h_{ij}(t) = H_{ij} e^{(-i\Omega t (1 + \bar{n} \cdot \bar{q} ) + \Phi_0))}
\end{equation}
The amplitude tensor  $H_{ij}$ is
\begin{equation}
	H_{ij} = h_+ e_{ij}^+(\bar{n},\psi) + h_{\times} e_{ij}^{\times}(\bar{n},\psi)
\end{equation}
where $h_{+,\times}$ are the constant amplitudes of the gravitational plane wave, and $e_{ij}^{+, \times}(\bar{n}, \psi)$ are the polarisation tensors which are uniquely defined by the principal axes of the GW. $\psi$ is the polarisation angle of the GW. \newline

@@

\section{Kalman filtering}
